Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan

### 2019-08-26 analysis/paper/paper.Rmd
some edits based on committee's suggestions

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  In East Asia, direct European colonial rule on local communities was much less pronounced than in other parts of the world. In this region, the question of long-lasting indirect impacts on indigenous peoples in the periphery of colonial control remains largely unanswered. Using trade ornaments recovered from Kiwulan (1400-1900 AD) in northeastern Taiwan, I explore the indirect impacts of European colonial activities on local societies. I examine the cultural contexts, variety, and spatial distributions of trade ornaments in three periods relating to different cultural interactions. The results indicate the diversity of ornaments was greater during the European presence period compared to previous periods and their spatial distribution was clustered. This hints at increasing social inequality in a colonial context. This study demonstrates the use of foreign ornaments as a proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonization; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is physically isolated and there are no traces of European settlement. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. This makes Yilan an especially unique location for exploring \indirect colonial effects on local indigenous settlements because of isolation.

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period (L5 and L6), European period (L4), and Chinese period (L2 and L1).

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole areas, 3173 pieces from 30 burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments in the form of bead strings that consisted of hundreds to thousands of pieces. In this paper I focus on the post hole areas where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organization. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

# Reproducibility and open source materials 
To enable re-use of materials and improve reprodicibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in XXX.XXX. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

```{r plot-all-densities, fig.cap= "Kernal density map for ornaments by periods"}
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)
png(
  here("analysis", "figures",
       "plot-kde-maps.png"), 
    width = 190,
    height = 63,
    units = "mm",
  res = 300)
par(mfrow = c(1, 3),
    oma = c(1,1,1,1) + 0.1,
    mar = c(0,0,1,1) + 0.1)
kernels_for_plotting <- 
  map2(ornaments_period_join_cen, 
       titles,
      ~density_analysis(.x, main = .y))
invisible(dev.off())
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-kde-maps.png"))
```

@@

Kernel density estimations (KDE) estimate the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here we use KDE to visualize core areas of ornaments and surrounding neighborhoods [@Bonnier2019; @Cortegoso2016]. Density values of artifacts per square meter were calculated for each cell.

The results show that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period (Figure \@ref(fig:plot-all-densities)). Those maps present three consistent sub-regions with a core area that shifts over time. The distribution might indicate different social groups who possessed more ornaments. The multiple groups during European period might reflect more unequal consumption of ornaments across the site, relative to other periods. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create an obvious hot spot. To further evaluate the pattern, we test the hypothesis that the distribution of ornaments is not randomly distributed to determine if social processes might explain ornament distributions.

@@

To test this hypothesis about the randomness of spatial locations, I used a Monte Carlo method to simulate average nearest-neighbor distances (ANN). Our null hypothesis is that the distribution of ornaments is consistent with a completely random process. We simulated the locations of ornaments across the space 1000 times for each time period. The observed data was compared to the random patterns generated by simulated processes to determine if our data are randomly distributed or not. Figure \@ref(fig:hypo-test-distribution) shows the distributions of the ANN distances calculated on 1000 simulations of ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European period, which means the ornaments have non-random clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese period, with about one third of the simulated values are greater than our observed ANN value. This testing reveals that the clustered distributions of ornaments during the pre-European and European periods might reflect the presence of different social groups. Moreover, the clustering of ornaments during the European period is highly non-random, potentially indicating that some of the social changes leading to a concentration of power to control the distribution of ornaments at Kiwulan.

@@

Examination of the indirect influence of colonialism can reveal how peripheral areas were influenced by the colonial activities or involved in the colonial economy centered in the major European colonies [@Trabert2017]. Kiwulan in Northeastern Taiwan is an important case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan can provide valuable insights into the discussion of indirect colonial influence. The frequency and spatial distribution of body ornaments at Kiwulan present three distinct patterns during different culture interaction periods. The greater amount and diversity of ornament types during the European period reflects international exchanges in a colonial context. Those ornaments may carry an exotic and powerful image that signals wealth and trading connections for the inhabitants of Kiwulan. These symbolic values may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality.

This study demonstrates that foreign ornaments can be a proxy to detect indirect colonial influence on local indigenous populations. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system. Future work could extend this approach to studies of other trade goods such as ceramics. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and future tests of this should include analysis of pottery production and standardization, and mortuary practices.

@@

I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the ornaments used in this study. I thank Shui-jin Chiu, Shu-hui Jian, Jiun-yao  Lai, and the staff in archaeology lab for their invaluable assistance in preparing samples for analysis. I also thank Yu-pei Chen for the discussion regarding the site at the earliest stage of fieldwork. This work was supported in part by the Fritz Graduate Fellowship and travel grants from the Department of Anthropology at University of Washington. I would also like to thank Ben Fitzhugh and Peter Lape, and anonymous reviewers for their insightful comments on early drafts. I especially thank Ben Marwick for providing academic support during this project, helping me with polishing figures with R langauge, and providing feedback on early drafts of this manuscript.

>>>>

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  In East Asia, direct European colonial rule on local communities was much less conspicuous than in other parts of the world. In this region, the question of long-lasting indirect impacts on indigenous peoples in the periphery of colonial control remains largely unanswered. Using trade ornaments recovered from Kiwulan (1400-1900 AD) in northeastern Taiwan, I explore the indirect impacts of European colonial activities on local societies. I examine the cultural contexts, variety, and spatial distributions of trade ornaments in three periods relating to different cultural interactions. The results indicate the diversity of ornaments was greater during the European presence period compared to previous periods and their spatial distribution was clustered. This hints at increasing social inequality in a colonial context. This study demonstrates the use of foreign ornaments as a proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonization; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is a physically isolated area adjacent to colonial centers and there are no traces of European settlement but occasional interactions with the Europeans. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. This makes Yilan an especially unique location for exploring indirect colonial effects on local indigenous settlements living beyond the borders of direct colonial occupation.

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period (L5 and L6), European period (L4), and Chinese period (L2 and L1) to indicate the dominant foreign influence and cultural interaction for different time periods.

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole areas, 3173 pieces from 30 burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments in the form of bead strings that consisted of hundreds to thousands of pieces. The ornaments from the burial context are not analyzed here since not many burials can be assigned to a time period with confidence. In this paper I focus on the post hole areas where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organization. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

# Reproducibility and open source materials

To enable re-use of materials and improve reprodicibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in XXX.XXX. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

```{r plot-all-densities, fig.cap= "Kernal density map for ornaments by periods. Used the bandwidth based on Silverman(1986)'s rule of thumb."}
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)
png(
  here("analysis", "figures",
       "plot-kde-maps.png"), 
    width = 190,
    height = 63,
    units = "mm",
  res = 300)
par(mfrow = c(1, 3),
    oma = c(1,1,1,1) + 0.1,
    mar = c(0,0,1,1) + 0.1)
kernels_for_plotting <- 
  map2(ornaments_period_join_cen, 
       titles,
      ~density_analysis(.x, main = .y))
invisible(dev.off())
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-kde-maps.png"))
```

@@

Kernel density estimations (KDE) estimate the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here I use KDE to visualize core areas of ornaments and surrounding neighborhoods [@Bonnier2019; @Cortegoso2016]. Density values of artifacts per square meter were calculated for each cell.

The results show that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period according to the kernal density maps (Figure \@ref(fig:plot-all-densities)). It seems that there are three consistent sub-regions with a core area that shifts over time. The distribution might indicate different household-based social groups who possessed more ornaments. The multiple groups during European period might reflect more unequal consumption of ornaments across the site, relative to other periods, to indicate social processes caused by human behavior. However, it could also just a random pattern resulted from bigger sample size. In addition, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create an obvious hot spot. To further evaluate the pattern, we test the hypothesis that the distribution of ornaments is not randomly distributed to determine if social processes might explain ornament distributions.

@@

To test this hypothesis about the randomness of spatial locations, I used a Monte Carlo method to simulate average nearest-neighbor distances (ANN). Our null hypothesis is that the distribution of ornaments is consistent with a completely random process. We simulated the locations of ornaments across the space 1000 times for each time period. The observed data was compared to the random patterns generated by simulated processes to determine if our data are randomly distributed or not. Figure \@ref(fig:hypo-test-distribution) shows the distributions of the ANN distances calculated on 1000 simulations of ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European period, which means the ornaments have non-random clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese period, with about one third of the simulated values are greater than our observed ANN value. This testing reveals that the clustered distributions of ornaments during the pre-European and European periods might reflect the presence of different social groups. Moreover, the clustering of ornaments during the European period is highly non-random, might potentially indicating that some of the social changes leading to different degrees of access foreign ornaments or a concentration of power to control the distribution of ornaments at Kiwulan.

@@

Examination of the indirect influence of colonialism can reveal how peripheral areas were influenced by the colonial activities or involved in the colonial economy centered in the major European colonies [@Trabert2017]. Kiwulan in Northeastern Taiwan is an important case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan can provide valuable insights into the discussion of indirect colonial influence on the local societies living beyond the borders of direct European colonial occupation. The frequency and spatial distribution of body ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects international exchanges in a colonial context. Ornaments were viewed as prestige goods in local indigenous culture before European contact. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may be intensified that further signals wealth and privileged trading connections with European colonists for the inhabitants of Kiwulan. These symbolic values may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality.

This study demonstrates that foreign ornaments can be a proxy to detect indirect colonial influence on local indigenous populations. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system and use them in their daily lives to display or intensify status differences. Future work could extend this approach to studies of other trade goods such as ceramics. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and future tests of this should include analysis of pottery production and standardization, and mortuary practices.

@@

I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the ornaments used in this study. I thank Shui-jin Chiu, Shu-hui Jian, Jiun-yao  Lai, and the staff in archaeology lab for their invaluable assistance in preparing samples for analysis. I also thank Yu-pei Chen for the discussion regarding the site at the earliest stage of fieldwork. This work was supported in part by the Fritz Graduate Fellowship and travel grants from the Department of Anthropology at University of Washington. I would also like to thank Ben Fitzhugh and Peter Lape for their insightful comments on the early draft. I especially thank Ben Marwick for providing academic support during this project, helping me with polishing figures with R langauge, and providing feedback on early drafts of this manuscript.
